Concentration Inequalities

### 2015-05-20 causal-bandit/importance_sampling.tex
lots of changes - laptop has battery issues
prior to re-write of observeThenBest approach

\title{Estimators}

\begin{document}
\def\ci{\perp\!\!\!\perp}
\maketitle

Suppose we have samples $(X_t,Y_t)$ for $t = 1... n$ drawn from the joint distribution $P(X,Y)$ and we want to estimate $\mu = P(Y=1|X=1)$

\section{The standard estimator}

\eqn{
\hat{\mu} = \frac{\sum_{t=1}^n Y_t \ind{X_t=1}}{\sum_{t=1}^n\ind{X_t=1}}
}

We can't actually say that this estimator is unbiased - as it can be undefined if $n = 0$

\color{red}{Is there an unbiased estimator for this simple problem if the $P(X = 1)$ is not known?}\color{black}

Let's consider variance and bounds.

>>>>

\title{Concentration Inequalities}

\begin{document}
\def\ci{\perp\!\!\!\perp}

\section{Hoeffding's Inequality}
If $X_1...X_n$ are independent, bounded random variables with $a_i < X_i < b_i$ and means $\mu_1...\mu_n$, and let $\mu = \frac{1}{n} \sum_{i=1}^n \mu_i$

\eqn {
P(\frac{1}{n}\sum_{i=1}^n X_i - \mu] \geq \epsilon) \leq \exp{\frac{-2n^2\epsilon^2}{\sum_{i=1}^n(b_i - a_i)^2}}
}

\section{Bernstein's Inequality}
$X_1...X_N$ are independent, bounded random variables with $0 \leq X_i \leq E[X_i]+b$ Let $V = \sum[V[X_i]$

\eqn{
&P(\sum X_i - \sum E[X_i] \geq \epsilon) \leq e^{-\frac{\epsilon^2}{2\sum V[X_i]+2b\epsilon/3}} \implies P\left(\boldsymbol{.} \geq \frac{2b\log(1/\delta)}{3})+\sqrt{2\log(1/\delta)\sum V[X_i]}\right)\leq \delta\\
&P(\frac{\sum X_i}{n} - \frac{\sum E[X_i]}{n} \geq \epsilon) \leq e^{-\frac{n\epsilon^2}{2\frac{1}{n}\sum V[X_i]+2b\epsilon/3}} \implies P\left(\boldsymbol{.} \geq \frac{2b\log(1/\delta)}{3n})+\frac{\sqrt{2\log(1/\delta)\sum V[X_i]}}{n}\right)\leq \delta
}

\subsection*{Estimating bounds for sums of bernoulli RV's with random denominator}
Suppose we have samples $(X_t,Y_t)$ for $t = 1... n$ drawn from the joint distribution $P(X,Y)$, where $X$ and $Y$ are binary random variables and $P(X) = p$ and we want to estimate $\mu = P(Y=1|X=1)$

\subsubsection*{Importance sampling estimate}
 Let: 
\eqn{
&\hat{\mu} = \frac{1}{n}\sum_{t=1}^nZ_t, \text{ where }
Z_t = \frac{Y_t \ind{X_t=1}}{p} \in \{0,\frac{1}{p}\}\\
&W_t = pZ_t, \; W_t \in \{0,1\},\; V[W_t] \leq p
}

\eqn{
&\text{Hoeff. } \implies P(\hat{\mu} - \mu \geq \epsilon) \leq e^{-2np^2\epsilon^2} 
\implies P(\hat{\mu} - \mu \geq \frac{1}{p}\sqrt{\frac{1}{2n}\log{\frac{1}{\delta}}}) \leq \delta \\
&\text{Berns. } \implies P(\mu -\hat{\mu}\geq \epsilon) \leq e^{-\frac{np\epsilon^2}{2(1+\epsilon/3)}} 
\implies P(\mu - \hat{\mu}\geq \frac{2\log(1/\delta)}{3np}+\sqrt{\frac{2}{np}\log{\frac{1}{\delta}}}) \leq \delta
}

\subsubsection*{Standard Estimator}
Let 
\eqn{
\hat{\mu} = \frac{\sum_{t=1}^n Y_t \ind{X_t=1}}{\sum_{t=1}^n\ind{X_t=1}}
}

We have a problem if we don't have a lower bound on $\sum_{t=1}^n\ind{X_t=1}$ as the estimator may be undefined. Assume $s = \sum_{t=1}^n\ind{X_t=1} > 0$, then:

\eqn{
\text{Hoeff.} \implies P(\hat{\mu} - \mu \geq \epsilon) \leq e^{-2s\epsilon^2}
 \implies P\left(\hat{\mu} - \mu \geq \sqrt{\frac{1}{2s}\log(\frac{1}{\delta})}\right)
}

### 2015-05-20 causal-bandit/observe_then_best.tex
prior to re-write of observeThenBest approach

@@

>>>>

\subsection{Generalizing to unknown q}
\begin{itemize}
\item Assume without loss of generality that $q_i \in [0,\frac{1}{2}]$

\item Assume we observe for the first $h/2$ timesteps.

\item Let $\hat{q}_i = \frac{2}{h}\sum_{t=1}^{h/2}\ind{X_i=1}$

\item Let $\bar{q}_i = min(\hat{q},1-\hat{q})$ We order the $\bar{q}$'s from smallest to largest and estimate $m$ with $\hat{m} = min_j : \bar{q}_j \geq \frac{1}{j}$

\item The worst case $\boldsymbol{q}$ for a given fixed $m$ is 
\eqn{
q_i = \begin{cases}
0 & \text{ if } i < m\\
\frac{1}{m} & \text{ if } i \geq m
\end{cases}
}

In this case $\bar{q}_0 ... \bar{q}_{m-1} = 0$

For $i \geq m$
\eq{
P(q_i - \bar{q}_i \geq C) = P(q_i - \hat{q}_i \geq C|\hat{q}_i\leq \frac{1}{2})P(\hat{q}_i \leq \frac{1}{2})+P(q_i - (1-\hat{q}_i) \geq C | \hat{q}_i > \frac{1}{2})P(\hat{q}_i>\frac{1}{2})
}

The first term will dominate for all but very small $m$. Lets just work with that one for now.

Via Bernstein's Inequality
\eqn{
P(q_i - \hat{q}_i \geq C) \leq \exp(-\frac{hC^2}{4q_i+4C/3}) := \gamma \\
\implies P(q_i - \hat{q}_i \geq \frac{4\log(1/\gamma)}{3h}+\sqrt{\frac{4q_i\log(1/\gamma)}{h}})\leq \gamma
}

Actually I think I can use the Lower tail version  of Bernsteins

\eqn{
P(q_i - \hat{q}_i \geq C) \leq \exp(-\frac{hC^2}{4q_i}) := \gamma \\
\implies P(q_i - \hat{q}_i \geq 2\sqrt{\frac{\log(1/\gamma)}{mh}})\leq \gamma
}

Define $\hat{m}_i = \frac{1}{\hat{q}_i}$.

\eq{
\hat{q}_i \geq q_i - C \implies \hat{m}_i \leq \frac{1}{q_i-C} = \frac{1}{\frac{1}{m} - C} \text{, where } C < \frac{1}{m} \\
\implies q_i - \hat{q}_i \leq C \implies \hat{m}_i - m \leq \frac{m^2C}{1-mC}
}

If we let $\gamma =  \frac{1}{N}\exp(-{\frac{h}{16m}})$

\eqn {
C = 2\sqrt{\frac{m\log N}{hm^2}+\frac{1}{16m^2}} \leq \frac{1}{2m} \text{, if } h > 16m\log N \\
}

Then

\eqn{
 q_i - \hat{q}_i \leq \frac{1}{2m} \implies \hat{m}_i - m \leq m \\
 \implies P(\hat{m}_i - m \geq m) \leq \frac{1}{N}\exp(-{\frac{h}{16m}})
}

If $\hat{m}_i - m \geq \varphi$ for at most $\varphi$ of the variables $i \geq m$, then $\hat{m} - m \leq \varphi$

Let $W_i = \ind{\hat{m_i}-m \geq \varphi}$, $E[W_i] \leq \gamma$,$V[W_i] \leq \gamma$ 
\eqn{
P(\sum_{i=m}^N W_i \geq (N-m)\gamma + \varepsilon) \leq \exp(-\frac{\varepsilon^2}{2(N-m)\gamma+2\varepsilon/3})
}
Letting $(N-m)\gamma + \varepsilon = \varphi \implies \varepsilon = \varphi - (N-m)\gamma$
\eqn{
P(\hat{m}-m \geq \varphi) =  P(\sum_{i=m}^N W_i \geq \varphi) &\leq \exp({-\frac{(\varphi-(N-m)\gamma)^2}{2(N-m)\gamma + 2(\varphi-(N-m)\gamma)/3}})\\
& \leq \exp({-\frac{(\varphi-N\gamma)^2}{2N\gamma + 2\varphi/3}}) \text{, where } N\gamma < \varphi
}

If $P(\hat{m}-m \geq \varphi) \leq \zeta$.

\eqn{
P(\hat{\mu}_i-\mu_i \geq \epsilon) \leq & P(\hat{\mu}_i-\mu_i \geq \epsilon |\hat{m}-m \leq \varphi )P(\hat{m}-m \leq \varphi) \\
& + P(\hat{\mu}_i-\mu_i \geq \epsilon |\hat{m}-m \geq \beta)P(\hat{m}-m \geq \varphi)\\  \leq & e^{-\frac{h\epsilon^2}{m+\varphi}} + \zeta e^{-\frac{h \epsilon^2}{N}}
}

Putting it together

\eqn {
P(\hat{\mu}_i-\mu_i \geq \epsilon) \leq \exp({-\frac{h\epsilon^2}{m+\varphi}}) + \exp({-\frac{(\varphi-N\gamma)^2}{2N\gamma + 2\varphi/3}}) \exp({-\frac{h \epsilon^2}{N}}) := \delta
}
The first term will dominate if

\eqn{
\frac{(\varphi-N\gamma)^2}{2N\gamma + 2\varphi/3} \geq \frac{h\epsilon^2}{m+\varphi}
}

Setting them equal and solving for $\varphi$. Roughly,

\eqn{
\varphi = \frac{h\epsilon^2}{m+\varphi} \implies \varphi = \frac{1}{2}(\sqrt{4\epsilon^2 h+ m^2} - m)\\
\implies \epsilon = \sqrt{\frac{\log(1/\delta)(\log(1/\delta)+m)}{h}} \leq \frac{\log(1/\delta)}{\sqrt{h}} + \sqrt{\frac{m}{h}\log(1/\delta)}
}

Now repeating the above but more exactly ...

\eqn{
\frac{(\varphi-N\gamma)^2}{2N\gamma + 2\varphi/3}  \geq & \frac{\varphi^2-2\varphi N\gamma}{4\varphi/3} \; \text{, if } \varphi > 3N\gamma \\
= & \frac{3}{4}(\varphi - 2N\gamma)
}

\eqn{
\frac{3}{4}(\varphi - 2N\gamma) = \frac{h\epsilon^2}{m+\varphi} \implies \varphi = &\frac{\sqrt{3}}{6}\sqrt{16\epsilon^2 h+3m^2+12N\gamma(m+N\gamma)} + N\gamma - \frac{m}{2}\\
\leq & \frac{\sqrt{3}}{2}\sqrt{4\epsilon^2 h+3mN\gamma}
}

If we let $\varphi =\frac{\sqrt{3}}{6}\sqrt{16\epsilon^2 h+3m^2+12N\gamma(m+N\gamma)} + N\gamma - \frac{m}{2} $

\eqn{
P(\hat{\mu}_i-\mu_i \geq \epsilon) \leq 2\exp(-\frac{3}{4}(\varphi - 2N\gamma))\\
\implies \epsilon = \sqrt{\frac{4\log^2(2/\delta)+(3m+6N\gamma)\log(2/\delta)}{3h}}\\
\leq \frac{2\log(2/\delta)}{\sqrt{3h}}+\sqrt{\frac{m+2N\gamma}{h}\log(2/\delta)}
}
Ok - but I still need to ensure there exists a $\gamma$ such that  $N\gamma < \frac{\varphi}{3}$ and $P(\hat{m}_i - m > \varphi) \leq \gamma$

\eqn{
P(\hat{m}_i - m > a) \leq \exp(-\frac{h a^2}{4m(m+a)^2})
}

It feels like I have just gone round and round in circles and pushed the problem to somewhere else ... my expression for $\varphi$ now contains both $\epsilon$ and $\gamma$ ...

\eqn{
\varphi \geq \frac{1}{2}\sqrt{m^2+4N\gamma(m+N\gamma)} + N\gamma - \frac{m}{2}
}

\end{itemize}

@@

\eqn{
 \implies & C = \frac{\varphi}{m(\varphi+m)}  \geq \frac{\varphi}{\alpha m^2} \text{ if } \varphi < (\alpha-1)m\\
\implies & P\left(\hat{m}_i - m \geq \alpha m^2 \left( \frac{4\log(1/\gamma)}{3h}+\sqrt{\frac{4\log(1/\gamma)}{mh}} \right)\right) \leq \gamma
}

If we let $\gamma = \frac{e^{-h/m}}{N}$

\eqn{
P\left(\hat{m}_i - m \geq \alpha m^2 \left( \frac{4}{3}(\frac{\log N}{h}+\frac{1}{m})+2\sqrt{\frac{\log N}{mh}+\frac{1}{m^2}} \right)\right) \leq \gamma \\
\implies P\left(\hat{m}_i - m \geq \alpha m^2 \left( \frac{4}{3}(\frac{2}{m})+2\sqrt{\frac{2}{m^2}} \right)\right) \leq \gamma \; \text{, if } h > m\log N \\
\implies P\left(\hat{m}_i - m \geq 6\alpha m \right) \leq \gamma
}

This doesn't work. We need $\hat{m}_i - m < (\alpha-1)m$ in order to get this expression.

Or we could let $\varphi = \frac{\sqrt{3}}{2}\sqrt{4\epsilon^2 h+3mN\gamma}$

\eqn{
P(\hat{\mu}_i-\mu_i \geq \epsilon) \leq 2\exp({-\frac{h\epsilon^2}{m+\frac{\sqrt{3}}{2}\sqrt{4\epsilon^2 h+3mN\gamma}}})
}
\eqn{
\implies \epsilon = & \frac{1}{\sqrt{2}}\sqrt{\frac{3 \log^2(2/\delta)+2m\log(2/\delta)}{h}+\frac{\sqrt{3}}{h^2}\sqrt{h^2\log^2(2/\delta)(3\log^2(2/\delta)+4m\log(2/\delta)+3mN\gamma)}} \\
\leq & 
}
